Title Goes Here

### 2021-01-05 analysis/paper/paper.Rmd
Muchos cambios

guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorsteps(title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

@@

```{r}
eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
  # .[, hgt := Anomaly(hgt), by = .(season(time), lon, lat, lev)] %>%
  .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), by = lev]

@@

```{r}
plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {

@@

```{r, fig.cap = "Correlación entre el las anomalías zonales de geopotential de cada mes y el campo medio de anomalías zonales de geopotencial (negro) y correlación (inversa) entre anomalías zonales de altura geopotencial y la parte real del EOF1 en (rojo) entre 80S y 20S."}
# cor_mean <- era5 %>% 
#   .[lev %in% c(50, 200)] %>% 
#   .[lat <= -20] %>% 
#   # .[lat >= -80] %>% 
#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
#   .[, hgt_mean := mean(hgt), by = .(lat, lon, lev)] %>% 
#   .[, .(mean = cor(hgt_mean, hgt)), by = .(lev, time)] %>%
#   identity()
# 
# 
# ceof$eof[[1]] %>% 
#   .$right %>% 
# 
#   .[PC == "PC1"] %>% 
#   sep_ReIm(format = "wide") %>% 
#   .[, hgt := NULL] %>% 
#   .[, PC := NULL] %>% 
#   era5[., on = .NATURAL] %>% 
#   .[, hgt := Anomaly(hgt), by = .(time, lat, lev)] %>% 
#   .[, .(hgt = -cor(hgt, R)), by = .(lev, time)] %>% 
# 
#   ggplot(aes(time, hgt))  +
#   geom_line(color = "red") +
#   geom_line(data = cor_mean, aes(y = mean)) +
#   facet_grid(lev~.)

@@

```{r}
regr <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("R", "I")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept()
```

```{r}
plot_regr <- function(data, which_PC, which_levs, which_var) {
  data <- data %>%
    .[ PC %in% which_PC] %>%
    .[variable == which_var]

  if (length(which_PC) > 1) {
    facet <- ggh4x::facet_nested(PC ~  term, labeller = labeller(term = lab_cplx))
  } else {
    facet <- ggh4x::facet_nested(lev ~  term, labeller = labeller(term = lab_cplx,
                                                                  lev = lab_lev))
  }

  if ("lev" %in% colnames(data)) {
    data <- data[lev %in% which_levs]
  } else {
    data[, lev := which_levs]
  }
  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
        geom_contour_tanaka() +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom())
    ) +

@@

```{r, fig.cap = "Regresiones con altura geopotencial PC1"}
plot_regr(regr, "PC1", c(200, 50), "hgt")
```

```{r, fig.cap = "Regresiones con temperatura y PC1"}
plot_regr(regr, "PC1", c(200, 50), "t")
```

```{r, fig.cap = "Regresiones con altura geopotencial y PC2"}
plot_regr(regr, "PC2", c(200, 50), "hgt")
```

```{r, fig.cap = "Regresiones con temperatura y PC2"}
plot_regr(regr, "PC2", c(200, 50), "t")
```

```{r}
regr_simpler <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[era5, on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(200, 50)] %>%
  .[, FitLm(hgt, I, R, se = TRUE), by = .(lev, lon, lat, PC, season)] %>%
  rm_intercept()
```

```{r, fig.cap = "Varianza temporal de altura geopotencial explicada por la variación temporal de cada componente principal."}
regr_simpler %>%
  # .[term == "R"] %>%
  .[lat <= -20] %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +

  geom_contour_fill(aes(z = r.squared, fill = ..level..),
                    breaks = MakeBreaks(0.1, exclude = 0)) +
  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom())  +

  geom_qmap(~.x[lat <= -20]) +
  scale_x_longitude() +
  scale_y_latitude() +
  ggh4x::facet_nested(lev ~ PC, labeller = labeller(lev = lab_lev)) +
  coord_polar() +
  axis_labs_smol
```

```{r, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
regr %>%
  .[lat <= -20] %>%
  .[lev %in% c(200, 50)] %>%
  # .[PC == "PC1"] %>%
  .[, FitWave(estimate, 1:3), by = .(lev, PC, term, lat)] %>%
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  ggh4x::facet_nested(lev + PC ~ term, scales = "free") +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude() +
  coord_flip()
```

@@

```{r}
sam <- "/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  .[, .(time, sam = -estimate_norm, lev, symmetry = term)] %>%
  .[, sam := sam/sd(sam), by = .(lev, symmetry)] %>%
  .[season(time) == "SON"] %>%
  .[, .(sam = mean(sam)), by = .(symmetry, lev, time = seasonally(time))]

@@

  ceof[, eof[[1]]$left] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    index[., on = "time"] %>%
    na.omit() %>%
    .[, index := get(index_col)] %>%
    ggplot(aes(hgt, index)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
                                               x = -1.2,
                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
                                                                                       newdata = data.frame(hgt = -1.2))),
                                           by = .(PC, part)],
                                bg.colour = "white", bg.r = 0.2,
                                size = 5,
                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
    scale_x_continuous("EOF") +
    scale_y_continuous(toupper(index_col)) +
    scale_color_brewer("Part", palette = "Dark2") +
    facet_grid(part~PC)
}
```

```{r, fig.cap = "Relación del EOF (estandarizado) con SAM"}
plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
```

```{r}

@@

"/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  # .[lev %in% c(200, 50)] %>%
  # .[term != "full"] %>%
  rm_singleton() %>%
  .[season(time) == "SON"] %>%
  .[, .(estimate_norm = mean(estimate_norm)), by = .(lev, term, time = seasonally(time))] %>%
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate_norm)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC) +
  coord_flip()

@@

```{r, fig.cap = "Idem anterior pero con ENSO"}
plot_with_index(ceof, enso)
```

@@

```{r}
psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
  .[, value := NULL]
```

```{r}
regr_psi <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, R, I, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
  rm_intercept() %>%
  .[, psi.z := estimate] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)]
```

```{r}

plot_psi <- function(data, which_pc, which_lev) {
  data %>%
    .[PC == which_pc] %>%
    .[lev == which_lev] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
                    size = 0.2, res = 2) +
    geom_qmap(keep = .015) +
    scale_fill_divergent(guide = "none") +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
    coord_quickmap(ylim = c(NA, 10)) +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol
}

```{r}
sst <- ERA5_SST() %>%
  ReadNetCDF(vars = c("t" = "sst")) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(t = mean(t)), by = .(lon, lat, time = seasonally(time))]
```

```{r}
regr_sst <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
  rm_intercept()
```

```{r}
plot_sst <- function(data, which_pc) {
  data %>%
    .[PC == which_pc] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15))
    ) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_qmap(keep = .015) +

    scale_x_longitude() +
    scale_y_latitude() +
    facet_grid(term ~ PC, labeller = labeller(term = lab_cplx)) +
    axis_labs_smol +
    coord_quickmap(ylim = c(-90, 10))
}

```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_psi(regr_psi, "PC1", 200) +
  plot_sst(regr_sst, "PC1")
```

```{r, fig.cap = "Regresión con PSI en 200 hPa y sst, PC2"}
plot_psi(regr_psi, "PC2", 200) +
  plot_sst(regr_sst, "PC2")
```

@@

```{r}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, .(pp = mean(pp, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
```

```{r}
regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
  .[season == "SON"] %>%
  .[, FitLm(value, I, R, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
  rm_intercept()
```

```{r}

continents <- list(africa  = list(lon = c(0, 60),
                                  lat = c(-40, 0)),
                   oceania = list(lon = c(100, 180),
                                  lat = c(-50, -5)),
                   america = list(lon = c(265, 325),
                                  lat = c(-60, -10)))

add_continent <- function(dt) {
  dt[, continent := fcase(lon %between% continents$africa$lon &
                            lat %between% continents$africa$lat, "africa",
                          lon %between% continents$oceania$lon &
                            lat %between% continents$oceania$lat, "oceania",
                          lon %between% continents$america$lon &
                            lat %between% continents$america$lat, "america")] %>%
    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
    .[]
}

```{r}
plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent) {
  data %>%
    .[PC %in% which_pc] %>%
    .[term %in% which_term] %>%
    .[, lev := "sfc"] %>%
    add_continent() %>%
    .[continent == which_continent] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
        geom_contour_tanaka() +
        scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15))
    ) +

    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05),
                geom = "point", size = 1) +
    geom_qmap(~.x %>% add_continent() %>% .[continent == which_continent], keep = 1) +

@@

```{r, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, "PC1", which_continent =  "america")  +
  plot_pp(regr_pp, "PC2", which_continent =  "america")
```

```{r, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, "PC1", which_continent = "africa") +
  plot_pp(regr_pp, "PC2", which_continent = "africa")
```

>>>>

guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = FALSE,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

```{r test-irlba}
has_irlba <- requireNamespace("rilba", quietly = TRUE)
if (has_irlba) {
  stop("If installed, metR::EOF uses irlba, but that package has a bug with complex matrices\n(see: https://github.com/bwlewis/irlba/issues/55). Uninstall irlba with renv::remove(\"irlba\") and try again.")
}

@@

```{r compute-eof-naive}
eof_naive <- era5[lev %in% c(50, 200) & lat %between% c(-90, -20)] %>% 
  # .[, hgt := Anomaly(hgt), by = .(season(time), lon, lat, lev)] %>%
  .[, hgt  := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(time, lev, lat)] %>% 
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1:4, data = .SD, suffix = "EOF"))), by = lev]

@@

```{r define-plot_ceof}
plot_ceof <- function(data, n = 1:2, formula = lev  ~  PC) {

@@

@@

```{r compute-regression}
regr <- ceof[, eof[[1]]$left, by = season] %>%
  .[season == "SON"] %>%
  sep_ReIm(format = "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = as.numeric(scale(R)), I = as.numeric(scale(I))), by = .(season, PC)] %>%
  dcast(time ~ PC, value.var = c("R", "I")) %>%
  .[melt(era5, id.vars = c("time", "lon", "lat", "lev")), on = "time", allow.cartesian = TRUE] %>%
  .[lev %in% c(850, 700, 200, 50)] %>%
  # .[variable == "hgt"] %>%
  .[, FitLm(value, I_PC1, R_PC1, I_PC2, R_PC2, se = TRUE), by = .(lev, lon, lat, variable)] %>%
  .[, c("term", "PC") := tstrsplit(term, "_")] %>%
  rm_intercept() %>% 
  .[, term := factor(term, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
```

```{r define-plot_regr}
plot_regr <- function(data, which_PC, which_levs, which_var) {
  data <- data %>%
    .[ PC %in% which_PC] %>%
    .[variable == which_var]

  var_title <- switch(which_var, 
                      hgt = "Geopotential height", 
                      t = "Temperature")
  lab_pc <- c(PC1 = "EOF1", PC2 = "EOF2")

  if (length(which_PC) > 1) {
    facet <- ggh4x::facet_nested(PC ~  term)
  } else {
    facet <- ggh4x::facet_nested(lev ~ PC +  term, labeller = labeller(lev = lab_lev, PC = lab_pc))
  }

  if ("lev" %in% colnames(data)) {
    data <- data[lev %in% which_levs]
  } else {
    data[, lev := which_levs]
  }
  data[lat <= -20] %>%
    # .[term == "I" & lev == 50] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    ditto(
      geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(exclude = 0), global.breaks = FALSE) +
        geom_contour_tanaka() +
        scale_fill_divergent(var_title, guide = guide_colorstrip_bottom())
    ) +

@@

(ref:eof1-regr-gh-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON geopotential height for the 1979 -- 2019 period. These coefficients come from multiple linear regression involving the real and imaginary part of both EFO2.

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt")
```

(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.

```{r eof1-regr-t, fig.cap = "(ref:eof1-regr-t-cap)"}
plot_regr(regr, "PC1", c(200, 50), "t") 
```

(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF.

```{r eof2-regr-gh, fig.cap = "(ref:eof2-regr-gh-cap)"}
plot_regr(regr, "PC2", c(200, 50), "hgt")
```

(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.

```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
plot_regr(regr, "PC2", c(200, 50), "t")
```

```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
regr %>%
  .[lat <= -20] %>%
  .[lev %in% c(200, 50)] %>%
  # .[PC == "PC1"] %>%
  .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  ggh4x::facet_nested(lev + PC ~ term, scales = "free") +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude() +
  coord_flip()
```

@@

```{r read-indices}
sam <- "/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  .[, .(time, sam = -estimate_norm, lev, symmetry = term)] %>%
  .[, sam := sam/sd(sam), by = .(lev, symmetry)] %>%
  .[season(time) == "SON"] %>%
  .[, .(sam = mean(sam)), by = .(symmetry, lev, time = seasonally(time))]

@@

  ceof[, eof[[1]]$left] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    index[., on = "time"] %>%
    na.omit() %>%
    .[, index := get(index_col)] %>%
    ggplot(aes(hgt, index)) +
    geom_point(aes(color = part)) +
    geom_smooth(aes(color = part), method = "lm") +
    shadowtext::geom_shadowtext(data = ~.x[, .(cor = cor(index, hgt),
                                               x = -1.2,
                                               y = sign(-cor(index, hgt))*.1 + predict(lm(index ~ hgt),
                                                                                       newdata = data.frame(hgt = -1.2))),
                                           by = .(PC, part)],
                                bg.colour = "white", bg.r = 0.2,
                                size = 5,
                                aes(x, y, color = part, label = paste0("r² = ", round(cor^2, 2)))) +
    scale_x_continuous("EOF") +
    scale_y_continuous(toupper(index_col)) +
    scale_color_brewer("Part", palette = "Dark2") +
    facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2")))
}
```

```{r sam-eof, fig.cap = "Standarised temporal EOFs and SAM index computed as the leading EOF of geopotential height anomalies at 700 hPa."}
plot_with_index(ceof, sam[lev == 700 & symmetry == "full"][, .(time, sam)])
```

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM and Symmetric SAM indices computed at each level according to Campitelli on press??.

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}

@@

"/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  # .[lev %in% c(200, 50)] %>%
  # .[term != "full"] %>%
  rm_singleton() %>%
  .[season(time) == "SON"] %>%
  .[, .(estimate_norm = mean(estimate_norm)), by = .(lev, term, time = seasonally(time))] %>%
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate_norm)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip()

@@

(ref:oni-eof-cap) Same as Figure \@ref(fig:sam-eof) but for the Oceanic Enso Index (REF??).

```{r oni-eof, fig.cap = "(ref:oni-eof-cap)"}
plot_with_index(ceof, enso)
```

@@

```{r compute-psi}
psi <- ReadNetCDF(ERA5(), vars = "vo",
                  subset = list(level = c(50, 200),
                                time = c("1979-01-01", "2019-12-01"))) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(vo = mean(vo)), by = .(lon, lat, lev,time = seasonally(time))] %>%
  .[, solve_poisson(vo, lon, lat), by = .(lev, time)] %>%
  .[, psi.z := Anomaly(value), by = .(time, lev, lat)] %>%
  .[, value := NULL]
```

```{r regr_psi}
regr_psi <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[psi, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(psi.z, R, I, se = TRUE),by = .(lon, lat, lev, PC, season)] %>%
  rm_intercept() %>%
  .[, psi.z := estimate] %>%
  .[, c("fx", "fy") := shceof::WaveFlux(.SD, p = 200), by = .(lev, PC, season, term)]
```

```{r sst}
sst <- ERA5_SST() %>%
  ReadNetCDF(vars = c("t" = "sst")) %>%
  normalise_coords() %>%
  .[season(time) == "SON"] %>%
  .[, .(t = mean(t)), by = .(lon, lat, time = seasonally(time))]
```

```{r regr_sst}
regr_sst <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[sst, on = "time", allow.cartesian = TRUE] %>%
  .[season(time) == "SON"] %>%
  .[, FitLm(t, I, R, se = TRUE), by = .(lon, lat, PC, season)] %>%
  rm_intercept()
```

```{r plot_sst_psi}
plot_sst_psi <- function(which_pc, which_lev) {
  sst_gdata <- regr_sst[PC == which_pc] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    .[, var := "SST"]
  regr_psi %>% 
    .[PC == which_pc] %>%
    .[lev == which_lev] %>%
    # .[abs(estimate) > 1, estimate := NA] %>%
    .[lat %between% c(-90, 10)] %>%
    .[, p_val := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>%
    .[, var := "Streamfunction"] %>% 
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    geom_contour_fill(aes(z = estimate*1e8, fill = ..level..),  breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
                    size = 0.2, res = 2) +
    geom_qmap(keep = .015) +
    scale_fill_divergent_discretised("Streamfunction", guide = "none") +
    ggnewscale::new_scale_fill() +

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(data = sst_gdata, 
                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(data = sst_gdata, 
                  aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(data = sst_gdata,
                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 2)),
                geom = "point", size = 0.2, alpha = 0.5) +
    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var, labeller = labeller(term = lab_cplx, 
                                               PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
    coord_quickmap(ylim = c(NA, 10)) +
    theme(legend.box = "horizontal") +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol
}
```

```{r sst-psi-1,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_sst_psi(which_pc = "PC1", which_lev = 200)
```

```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_sst_psi(which_pc = "PC2", which_lev = 200)
```

```{r mean-psi, fig.cap = "mean streamfunction"}
psi %>% 
  .[lev == 200, .(psi.z = -mean(psi.z)), by = .(lon, lat)] %>% 
  .[, c("fx", "fy") := WaveFlux(.SD)] %>% 
  .[lat <= 10] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = psi.z)) +
  geom_streamline(aes(dx = fx, dy = fy), min.L = 3, L = 10, n = 30,
                  size = 0.2, res = 2) +
  geom_qmap(keep = .015) +
  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(width = 15)) +
  scale_x_longitude() +
  scale_y_latitude()  +
  coord_quickmap(ylim = c(-90, 10))
```

@@

```{r cmap}
cmap <- ReadNetCDF(CMAP(), vars = c(pp = "precip"),
                   subset = list(lat = -90:10)) %>%
  normalise_coords() %>%
  .[, .(pp = mean(pp, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))]
```

```{r regr_pp}
regr_pp <- ceof[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt, "wide") %>%
  .[, hgt := NULL] %>%
  .[, `:=`(R = R/sd(R), I = I/sd(I)), by = .(season, PC)] %>%
  .[melt(cmap, id.vars = c("lon", "lat", "time")), on = "time", allow.cartesian = TRUE] %>%
  .[season == "SON"] %>%
  .[, FitLm(value, I, R, se = TRUE), by = .(lon, lat, PC, variable, season)] %>%
  rm_intercept()
```

```{r add_continents}
continents <- list(africa  = list(lon = c(0, 60),
                                  lat = c(-40, 0)),
                   oceania = list(lon = c(100, 180),
                                  lat = c(-50, -13)),
                   america = list(lon = c(265, 325),
                                  lat = c(-60, -10)))

add_continent <- function(dt) {
  dt[, continent := fcase(lon %between% continents$africa$lon &
                            lat %between% continents$africa$lat, "africa",
                          lon %between% continents$oceania$lon &
                            lat %between% continents$oceania$lat, "oceania",
                          lon %between% continents$america$lon &
                            lat %between% continents$america$lat, "america")] %>%
    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
    .[]
}
```

```{r plot_pp}
plot_pp <- function(data, which_pc, which_term = c("R", "I"), which_continent = NULL) {
  filter_continent <- function(data) {
    if (!is.null(which_continent)) {
    data[ continent == which_continent]  
    } else {
      data
    }

  }
  data %>%
    .[PC %in% which_pc] %>%
    .[term %in% which_term] %>%
    .[, lev := "sfc"] %>%
    .[lat >= -60] %>% 
    add_continent() %>%
    filter_continent() %>% 
    # .[abs(estimate) <= 2] %>% 
    .[, p_val := Pvaluate(estimate, std.error, df, "none"), by = .(lev, PC, term)] %>%
    periodic(lon = c(0, 360)) %>%
    ggplot(aes(lon, lat)) +
    # geom_raster(aes(fill = estimate)) +
    geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +

    # scale_fill_divergent() +
    scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
    geom_contour2(aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(aes(subset = p_val <= 0.05),
                geom = "point", size = 0.2) +
    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% add_continent() %>% filter_continent(),
              keep = 1) +

@@

```{r pp-america, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(data = regr_pp, which_pc= c("PC1", "PC2"), which_continent =  "america")  
```

```{r pp-africa, fig.cap = "Regresión de cada PC con la precipitación"}
plot_pp(regr_pp, c("PC1", "PC2"), which_continent = "africa") 
```
